A geometric framework for capturing high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-29 paper/main.tex
add labels and \namerefs for section references, replace \hyperref with \nameref

We asked participants in our study to complete brief multiple-choice quizzes before, between, and after watching two lecture videos from the Khan Academy \citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture video, entitled \textit{Four
Fundamental Forces}, discussed the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second,
entitled \textit{Birth of Stars}, provided an overview of our current
understanding of how stars form.
%(See \hyperref[subsec:experiment]{\textit{Experiment}} for more information on how these lectures were chosen.)
%Each multiple-choice quiz consisted of 13 questions that would enable us to test participants'
%knowledge about each individual lecture, as well as related content not
%specifically presented in either video ( see Tab.~\questions).
%
%We selected these lessons to be accessible to a broad audience (i.e., requiring minimal prerequisite knowledge to understand) and largely independent of each other (i.e., focused on sufficiently different material that understanding one did not require having seen the other), but conceptually related to each other (covering at least \textit{some} similar or overlapping content).
We selected these lessons to be (a) accessible
to a broad audience, i.e., requiring minimal prerequisite knowledge to understand;
(b) conceptually related to each other, i.e., covering at least \textit{some}
similar or overlapping content; and (c) largely independent of each other, i.e., focused
on sufficiently different material that understanding one did not require having seen
the other. The two videos we selected are introductory lectures that both
belong to Khan Academy's ``Cosmology and Astronomy'' course domain, but are taken from
different lecture series (``Scale of the Universe'' and ``Stars, Black Holes, and Galaxies''
for the first and second lectures, respectively).

@@

To study how participants' conceptual knowledge changed over the course of the experiment, 
we first sought to characterize the abstract concepts presented to them in each of the two 
lectures. We adapted an approach we developed in prior work \citep{HeusEtal21} to extract 
the latent themes from the lectures' contents using a topic model \citep{BleiEtal03}. Briefly, 
topic models take as input a collection of text documents and learn a set of ``topics'' (i.e., 
latent themes) from their contents. Once fit, a topic model can be used to transform arbitrary 
(potentially new) documents into sets of ``topic proportions,'' describing the weighted blend 
of learned topics reflected in their texts. We parsed automatically generated transcripts of 
the two lectures into overlapping sliding windows, and used these as documents to fit our 
model (Fig.~\ref{fig:sliding-windows}A; see 
\hyperref[subsec:topic-modeling]{\textit{Constructing text embeddings of multiple lectures 
and questions}}). The ($k$=15) topics

%To construct a corpus of text documents for the two lectures, we parsed their transcripts into overlapping sliding windows (Fig.~\ref{fig:sliding-windows}A; see \hyperref[subsec:topic-modeling]{\textit{Constructing text embeddings of multiple lectures and questions}})

We trained a topic model \citep{BleiEtal03} using sliding windows of text from the two
lectures' transcripts (see \hyperref[subsec:topic-modeling]{\textit{Constructing text embeddings of multiple
lectures and questions}}). We also used the same model (i.e., trained on the
videos' transcripts) to embed the text of each question in our pool. This
yielded, for each second of each video, and for each question, a single
15-dimensional topic vector---i.e., a coordinate in a text embedding space
(Fig.~\ref{fig:sliding-windows}). Intuitively, each dimension of the embedding
space corresponds to a ``theme'' or ``topic'' reflected in some part(s) of the
videos (Tab.~\topics), and the coordinates in embedding space denote the blend
of themes reflected by a particular excerpt of text (e.g., from part of a
video's transcript, from a question, etc.).

@@

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \textit{Estimating dynamic knowledge traces}).
Essentially, the estimated knowledge at the coordinate is given by the weighted
average proportion of quiz questions the participant answered correctly, where
the weights reflect how much each question is ``about'' the content at $x$.
When we apply this approach to estimate the participant's knowledge about the
content presented in each moment of each lecture, we can obtain a detailed
timecourse describing how much ``knowledge'' the participant has about any part
of the lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also
apply this approach separately for the questions from each quiz the
participants took throughout the experiment. From just 13 questions per quiz,
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1106 samples across the two
lectures).

@@

    \caption{\textbf{Estimating moment-by-moment knowledge acquisition.}
    \textbf{A. Moment-by-moment knowledge about the \textit{Four Fundamental
    Forces}}. Each trace displays the weighted proportion of correctly answered
    questions about the content reflected in each moment of the lecture (see
    \textit{Estimating dynamic knowledge traces}), using responses from one
    quiz (color). The traces are averaged across participants. \textbf{B.
    Average estimated knowledge about the \textbf{Four Fundamental Forces}.}
    Each bar displays the across-timepoint average knowledge, estimated using
    the responses to one quiz's questions. \textbf{C. Moment-by-moment
    knowledge about the \textit{Birth of Stars}}. The panel is in the same
    format as Panel A, but here the knowledge estimates are for the
    moment-by-moment content of the \textit{Birth of Stars} lecture. \textbf{D.
    Average estimated knowledge about the \textit{Birth of Stars}.} The panel
    is in the same format as Panel B, but here the knowledge estimates are for
    the content of the \textit{Birth of Stars} lecture. All panels: error
    ribbons and error bars denote 95\% confidence intervals, estimated across
    participants.}

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see \textit{Creating
    knowledge and learning map visualizations}). The topic trajectories of each
    lecture and the coordinates of each question are indicated by dotted lines
    and dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps are
    in the same general format as the knowledge maps in Panel A, but each
    coordinate in the learning maps indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' across the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word reflected by the blend of topics
    represented at the locations of the stars in the maps. The words' colors
    indicate how much each word is weighted on average across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\textit{Creating knowledge and learning map visualizations}). Next, we sampled
points evenly from a $100 \times 100$ grid of coordinates that evenly tiled a
rectangle enclosing the 2D projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10K sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

@@

\subsubsection*{Estimating dynamic knowledge traces}

@@

\subsubsection*{Creating knowledge and learning map visualizations}

>>>>

We asked participants in our study to complete brief multiple-choice quizzes before, between, and after watching two lecture videos from the Khan Academy \citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture video, entitled \textit{Four
Fundamental Forces}, discussed the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second,
entitled \textit{Birth of Stars}, provided an overview of our current
understanding of how stars form.
%(See \textit{\nameref{subsec:experiment}} for more information on how these lectures were chosen.)
%Each multiple-choice quiz consisted of 13 questions that would enable us to test participants'
%knowledge about each individual lecture, as well as related content not
%specifically presented in either video ( see Tab.~\questions).
%
%We selected these lessons to be accessible to a broad audience (i.e., requiring minimal prerequisite knowledge to understand) and largely independent of each other (i.e., focused on sufficiently different material that understanding one did not require having seen the other), but conceptually related to each other (covering at least \textit{some} similar or overlapping content).
We selected these lessons to be (a) accessible
to a broad audience, i.e., requiring minimal prerequisite knowledge to understand;
(b) conceptually related to each other, i.e., covering at least \textit{some}
similar or overlapping content; and (c) largely independent of each other, i.e., focused
on sufficiently different material that understanding one did not require having seen
the other. The two videos we selected are introductory lectures that both
belong to Khan Academy's ``Cosmology and Astronomy'' course domain, but are taken from
different lecture series (``Scale of the Universe'' and ``Stars, Black Holes, and Galaxies''
for the first and second lectures, respectively).

@@

To study how participants' conceptual knowledge changed over the course of the experiment, 
we first sought to characterize the abstract concepts presented to them in each of the two 
lectures. We adapted an approach we developed in prior work \citep{HeusEtal21} to extract 
the latent themes from the lectures' contents using a topic model \citep{BleiEtal03}. Briefly, 
topic models take as input a collection of text documents and learn a set of ``topics'' (i.e., 
latent themes) from their contents. Once fit, a topic model can be used to transform arbitrary 
(potentially new) documents into sets of ``topic proportions,'' describing the weighted blend 
of learned topics reflected in their texts. We parsed automatically generated transcripts of 
the two lectures into overlapping sliding windows, and used these as documents to fit our 
model (Fig.~\ref{fig:sliding-windows}A; see \textit{\nameref{subsec:topic-modeling}}).
The ($k$=15) topics

%To construct a corpus of text documents for the two lectures, we parsed their transcripts into overlapping sliding windows (Fig.~\ref{fig:sliding-windows}A; see \textit{\nameref{subsec:topic-modeling}})

We trained a topic model \citep{BleiEtal03} using sliding windows of text from the two
lectures' transcripts (see \textit{\nameref{subsec:topic-modeling}}). We also used the same model (i.e., trained on the
videos' transcripts) to embed the text of each question in our pool. This
yielded, for each second of each video, and for each question, a single
15-dimensional topic vector---i.e., a coordinate in a text embedding space
(Fig.~\ref{fig:sliding-windows}). Intuitively, each dimension of the embedding
space corresponds to a ``theme'' or ``topic'' reflected in some part(s) of the
videos (Tab.~\topics), and the coordinates in embedding space denote the blend
of themes reflected by a particular excerpt of text (e.g., from part of a
video's transcript, from a question, etc.).

@@

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \textit{\nameref{subsec:traces}}).
Essentially, the estimated knowledge at the coordinate is given by the weighted
average proportion of quiz questions the participant answered correctly, where
the weights reflect how much each question is ``about'' the content at $x$.
When we apply this approach to estimate the participant's knowledge about the
content presented in each moment of each lecture, we can obtain a detailed
timecourse describing how much ``knowledge'' the participant has about any part
of the lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also
apply this approach separately for the questions from each quiz the
participants took throughout the experiment. From just 13 questions per quiz,
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1106 samples across the two
lectures).

@@

    \caption{\textbf{Estimating moment-by-moment knowledge acquisition.}
    \textbf{A. Moment-by-moment knowledge about the \textit{Four Fundamental
    Forces}}. Each trace displays the weighted proportion of correctly answered
    questions about the content reflected in each moment of the lecture (see
    \textit{\nameref{subsec:traces}}), using responses from one
    quiz (color). The traces are averaged across participants. \textbf{B.
    Average estimated knowledge about the \textbf{Four Fundamental Forces}.}
    Each bar displays the across-timepoint average knowledge, estimated using
    the responses to one quiz's questions. \textbf{C. Moment-by-moment
    knowledge about the \textit{Birth of Stars}}. The panel is in the same
    format as Panel A, but here the knowledge estimates are for the
    moment-by-moment content of the \textit{Birth of Stars} lecture. \textbf{D.
    Average estimated knowledge about the \textit{Birth of Stars}.} The panel
    is in the same format as Panel B, but here the knowledge estimates are for
    the content of the \textit{Birth of Stars} lecture. All panels: error
    ribbons and error bars denote 95\% confidence intervals, estimated across
    participants.}

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see 
    \textit{\nameref{subsec:knowledge-maps}}). The topic trajectories of each
    lecture and the coordinates of each question are indicated by dotted lines
    and dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps are
    in the same general format as the knowledge maps in Panel A, but each
    coordinate in the learning maps indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' across the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word reflected by the blend of topics
    represented at the locations of the stars in the maps. The words' colors
    indicate how much each word is weighted on average across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\textit{\nameref{subsec:knowledge-maps}}). Next, we sampled
points evenly from a $100 \times 100$ grid of coordinates that evenly tiled a
rectangle enclosing the 2D projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10K sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

@@

\subsubsection*{Estimating dynamic knowledge traces}\label{subsec:traces}

@@

\subsubsection*{Creating knowledge and learning map visualizations}\label{subsec:knowledge-maps}
